### 2018-07-21 methods.tex
Results section, keff, spectrum finished

\begin{figure}[hbp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{view_serpent.png}
  \caption{Plan and elevation view of SERPENT 2 \gls{MSBR} model developed in this work.}
  \label{fig:serpent_plan_view}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h!]
        %\centering
        \caption{Summary of principal data for MSBR \cite{robertson_conceptual_1971}.}
        \begin{tabularx}{\textwidth}{ s  s}
        \hline
                Thermal capacity of reactor           & 2250 MW(t)
                \\ 
                Net electrical output                 & 1000 MW(e) 
                \\  
                Net thermal efficiency        & 44.4\%
                \\  
                Salt volume fraction in central core zone     & 0.13
                \\ 
                Salt volume fraction in outer core zone       & 0.37
                \\ 
                Fuel salt inventory (Zone I)                  & 8.2 m$^3$	
                \\ 
                Fuel salt inventory (Zone II)                 & 10.8 m$^3$	
                \\ 
                Fuel salt inventory (annulus)                 & 3.8 m$^3$	
                \\  
                Total fuel salt inventory                     & 48.7 m$^3$	
                \\ 
                Fissile mass in fuel salt                   & 1303.7 kg	
                \\ 
                Fuel salt components                  & 
                LiF-BeF$_2$-ThF$_4$-$^{233}$UF$_4$	
                \\  
                Fuel salt composition                 & 
                71.75-16-12-0.25 mole\%
                \\
                Fuel salt density                    & 
                3.35 g/cm$^3$
                \\ \hline
        \end{tabularx}
        \label{tab:msbr_tab}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Online reprocessing method}
Removing specific chemical elements from a molten salt is a complicated task that requires intelligent design (e.g., chemical separations equipment design, fuel salt flows to equipment) and has a considerable economic cost.

\subsection{The SaltProc modeling and simulation code}

The SaltProc tool \cite{andrei_rykhlevskii_arfc/saltproc:_201*} is designed to expand SERPENT 2
 depletion capabilities for modeling liquid-fueled \gls{MSR} for continuous reprocessing.
The Python package uses HDF5 \cite{the_hdf_group_hierarchial_1997} to store data, and the Nuclear Engineering Toolkit PyNE \cite{scopatz_pyne:_2012}
for SEPRENT output file parsing and nuclide naming. It is an open-source tool that uses a semi-continuous approach to 
simulate continuous feeds and removals in the \gls{MSR}.

The tool structure and capabilities of SaltProc are similar to ChemTriton tool for SCALE
developed in \gls{ORNL} \cite{powers_new_2013}. SaltProc is coupled with Monte Carlo SERPENT 2
software which allows to simulate online reprocessing for irregular full-core geometry with 
high level of fidelity.  The primary function of SaltProc is to manage material streams while
SERPENT 2 performs most of the computationally heavy work, namely neutron transport and depletion
calculations. Saltproc is defined as a 
python class, where each material stream is defined as a isotopic atomic density
vector variable. This allows tracking of time-sensitive material streams such as the
Pa-233 tank in the \gls{MSBR}. The user can define the reprocessing parameters, such as the 
reprocessing interval and removal efficiency.  In addition, SaltProc provides a set of functions 
for each stream: read and write isotopic data in/from database, separate out specific isotopes 
from stream with defined efficiency, feed in specific isotopes to stream, and maintain constant 
number density of specific nuclide in the core. These attributes and functions are crucial to 
simulate the operation of a complex, multi-zone, multi-fluid \gls{MSR} and are universal for 
myriad reactor systems.

Saltproc is currently in active development in Github (https://github.com/arfc/saltproc), and has unit tests and
Travis continuous integration for sustainable development. There is also documentation
generated through Sphinx document generator for ease of use. We plan to implement
user-defined reprocessing scheme definition, two-region \gls{MSR} modeling capabilities,
and decay modeling in future releases.

@@

@@

\begin{itemize}
    \item \texttt{core adensity before reproc} (number of isotopes X timesteps)
    \item \texttt{core adensity after reproc} (number of isotopes X timesteps)
    \item \texttt{Keff_BOC} (1 X timesteps)
    \item \texttt{Keff_EOC} (1 X timesteps)
    \item \texttt{Th tank adensity} (number of isotopes X timesteps)
    \item \texttt{iso codes} (number of isotopes X 1)
\end{itemize}

Liquid-fueled \gls{MSBR} design in this thesis focuses on the state of the core at an equilibrium condition, after fission products have
built up in the fuel salt during years if operation. Isotopic composition of the fuel salt continues change slightly even after decades
of operation, but the dominant nuclides that have significant impact to the neutronic behavior tend to reach an equilibrium concentration
(e.g., vary less than 1\% over several years). In contrast, from the startup of an \gls{MSBR} until the equilibrium condition,
the fuel salt composition undergoes significant changes (e.g., changes in fission products, minor actinides, and fissile
materials number density). During this period, the material feeds and removals should be optimized for the fastest \gls{MSBR} transition to an
equilibrium state. A faster transition simplifies the reactor operation because, at equilibrium, the fissile and fertile feed rates,
fission product removal rates, and corresponding core safety parameters are constant in time.

\begin{figure}[htp!] % replace 't' with 'b' to 
  \centering
  \vspace{-0.3em}
  \includegraphics[width=\textwidth]{saltproc_flowchart.pdf}
  \caption{Flow chart for the Saltproc python-based tools.}
  \vspace{-1.0em}
  \label{fig:saltproc_flow}
\end{figure}
\FloatBarrier

>>>>

\begin{figure}[hbp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{view_serpent.png}
  \caption{Plan and elevation views of SERPENT 2 \gls{MSBR} model developed in this work.}
  \label{fig:serpent_plan_view}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h!]
        %\centering
        \caption{Summary of principal data for MSBR \cite{robertson_conceptual_1971}.}
        \begin{tabularx}{\textwidth}{ s  s}
        \hline
                Thermal capacity of reactor           		& 2250 MW(t)
                \\ 
                Net electrical output                 		& 1000 MW(e) 
                \\  
                Net thermal efficiency        				& 44.4\%
                \\  
                Salt volume fraction in central zone I		& 0.13
                \\ 
                Salt volume fraction in outer zone II       & 0.37
                \\ 
                Fuel salt inventory (Zone I)                & 8.2 m$^3$	
                \\ 
                Fuel salt inventory (Zone II)               & 10.8 m$^3$	
                \\ 
                Fuel salt inventory (annulus)               & 3.8 m$^3$	
                \\  
                Total fuel salt inventory                   & 48.7 m$^3$	
                \\ 
                Fissile mass in fuel salt                   & 1303.7 kg	
                \\ 
                Fuel salt components                  & 
                LiF-BeF$_2$-ThF$_4$-$^{233}$UF$_4$	
                \\  
                Fuel salt composition                 & 
                71.75-16-12-0.25 mole\%
                \\
                Fuel salt density                    & 
                3.35 g/cm$^3$
                \\ \hline
        \end{tabularx}
        \label{tab:msbr_tab}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Online reprocessing method}
Removing specific chemical elements from a molten salt is a complicated task that requires intelligent design (e.g., chemical separations equipment design, fuel salt flows to equipment) and has a considerable economic cost. All liquid-fueled \gls{MSR} designs involve varying levels of online fuel processing. Minimally, volatile gaseous fission products (e.g. Kr, Xe) escape from the fuel salt during routine reactor operation and must be captured. Additional systems might be used to enhance removal of those elements. Most designs also call for the removal of noble and rare earth metals from the core since these metals act as neutron poisons. Some designs suggest a more complex list of elements to process (figure ~\ref{fig:periodic_tab}), including the temporary removal of protactinium from the salt or other regulation of the actinide inventory in the fuel salt \cite{ahmad_neutronics_2015}.

\begin{figure}[htp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{periodic_map.png}
  \caption{Processing options for \gls{MSR} fuels \cite{ahmad_neutronics_2015}.}
  \label{fig:periodic_tab}
\end{figure}

\subsubsection{Fuel material flows}
The $^{232}$Th in the fuel absorbs thermal neutrons and produces $^{233}$Pa which then decays into the fissile $^{233}$U. Furthermore, the \gls{MSBR} design requires online reprocessing to remove all poisons (e.g. $^{135}$Xe), noble metals, and gases (e.g. $^{75}$Se, $^{85}$Kr) every 20 seconds. Protactinium presents a challenge, since it has a large absorption cross section in the thermal energy spectrum. Accordingly, $^{233}$Pa is continuously removed from the fuel salt into a protactinium decay tank to allow $^{233}$Pa to decay to $^{233}$U without poisoning the reactor. The reactor reprocessing system is designed to separate $^{233}$Pa from the molten-salt fuel over 3 days, hold it while $^{233}$Pa decays into $^{233}$U, and return it back to the primary loop. This feature allows the reactor to avoid neutron losses to protactinium, keeps fission products to a very low level, and increases the efficiency of $^{233}$U breeding. Table~\ref{tab:reprocessing_list} summarizes full list of nuclides and the cycle times used for modeling salt treatment and separations \cite{robertson_conceptual_1971}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
        \centering
        \caption{The effective cycle times for protactinium and fission products removal (reproduced from \cite{robertson_conceptual_1971}).}
        \begin{tabularx}{\textwidth}{ x | s | x }
        \hline 
        Processing group & \qquad\qquad\qquad Nuclides & Cycle time (at full power) \\ \hline 
        Rare earths & Y, La, Ce, Pr, Nd, Pm, Sm, Gd & 50 days \\ 
        \qquad & Eu & 500 days \\ 
        Noble metals & Se, Nb, Mo, Tc, Ru, Rh, Pd, Ag, Sb, Te & 20 sec \\
        Seminoble metals & Zr, Cd, In, Sn & 200 days \\
        Gases & Kr, Xe & 20 sec \\ 
        Volatile fluorides & Br, I & 60 days \\
        Discard & Rb, Sr, Cs, Ba & 3435 days \\ 
        Salt discard & Th, Li, Be, F & 3435 days \\ 
        Protactinium & $^{233}$Pa & 3 days \\ 
        Higher nuclides & $^{237}$Np, $^{242}$Pu & 16 years \\  \hline
        \end{tabularx}
        %\footnotetext{Chemical processing plant and gas separation system removing chemical elements (not isotopes) only. Isotopes instead of elements listed because other isotopes are short-lived and might be ignored.}
        \label{tab:reprocessing_list}
\end{table}
Th removal rates vary among nuclides in this reactor concept which dictate the necessary resolution of depletion calculations. If the depletion time intervals are very short, an enormous number of depletion steps are required to obtain the equilibrium composition. On the other hand, if the depletion  calculation time interval is too long, the impact of short-lived fission products is not captured. To compromise, the time interval for depletion calculations in this model was selected as 3 days to correlate with the removal interval of $^{233}$Pa and $^{232}$Th was continuously added to maintain the initial mass fraction of $^{232}$Th.

\subsubsection{The SaltProc modeling and simulation code}
The SaltProc tool \cite{andrei_rykhlevskii_arfc/saltproc:_2018} is designed to expand SERPENT 2
 depletion capabilities for modeling liquid-fueled \gls{MSR} for continuous reprocessing.
The Python package uses HDF5 \cite{the_hdf_group_hierarchical_1997} to store data, and the Nuclear Engineering Toolkit PyNE \cite{scopatz_pyne:_2012}
for SEPRENT output file parsing and nuclide naming. It is an open-source tool that uses a semi-continuous approach to 
simulate continuous feeds and removals in the \gls{MSR}.

The tool structure and capabilities of SaltProc are similar to ChemTriton tool for SCALE
developed in \gls{ORNL} \cite{powers_new_2013}. SaltProc is coupled with Monte Carlo SERPENT 2
software which allows to simulate online reprocessing for irregular full-core geometry with 
high level of fidelity.  The primary function of SaltProc is to manage material streams while
SERPENT 2 performs most of the computationally heavy work, namely neutron transport and depletion
calculations. Saltproc is defined as a 
python class, where each material stream is defined as a isotopic atomic density
vector variable. This allows tracking of time-sensitive material streams such as the
$^{233}$Pa tank in the \gls{MSBR}. The user can define the reprocessing parameters, such as the 
reprocessing interval and removal efficiency.  In addition, SaltProc provides a set of functions 
for each stream: read and write isotopic data in/from database, separate out specific isotopes 
from stream with defined efficiency, feed in specific isotopes to stream, and maintain constant 
number density of specific nuclide in the core. These attributes and functions are crucial to 
simulate the operation of a complex, multi-zone, multi-fluid \gls{MSR} and are universal for 
myriad reactor systems.

Saltproc is currently in active development in Github (https://github.com/ arfc/saltproc), and has unit tests and Travis continuous integration for sustainable development. There is also documentation
generated through Sphinx document generator for ease of use. We plan to implement
user-defined reprocessing scheme definition, two-region \gls{MSR} modeling capabilities,
and decay modeling in future releases.

@@

\begin{figure}[htp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=0.75\textwidth]{saltproc_flowchart.pdf}
  \caption{Flow chart for the Saltproc python-based tools.}
  \label{fig:saltproc_flow}
\end{figure}

@@

\begin{itemize}
    \item \texttt{core adensity before reproc} (number of isotopes x timesteps)
    \item \texttt{core adensity after reproc} (number of isotopes x timesteps)
    \item \texttt{Keff\_BOC} (1 x timesteps)
    \item \texttt{Keff\_EOC} (1 x timesteps)
    \item \texttt{Th tank adensity} (number of isotopes x timesteps)
    \item \texttt{iso codes} (number of isotopes x 1)
\end{itemize}

Liquid-fueled \gls{MSBR} design in this work focuses on the state of the core at an equilibrium condition, after fission products have built up in the fuel salt during years if operation. Isotopic composition of the fuel salt continues change slightly even after decades of operation, but the dominant nuclides that have significant impact to the neutronic behavior tend to reach an equilibrium concentration (e.g., vary less than 1\% over several years). In contrast, from the startup of an \gls{MSBR} until the equilibrium condition, the fuel salt composition undergoes significant changes (e.g., changes in fission products, minor actinides, and fissile materials number density). During this period, the material feeds and removals should be optimized for the fastest \gls{MSBR} transition to an equilibrium state. A faster transition simplifies the reactor operation because, at equilibrium, the fissile and fertile feed rates, fission product removal rates, and corresponding core safety parameters are constant in time.

### 2018-07-21 results.tex
Results section, keff, spectrum finished

\section{Results}

>>>>

\section{Results}
The SaltProc online reprocessing simulation code is demonstrated in four applications: (1) analyzing  \gls{MSBR} neutronics and fuel cycle to find the equilibrium core composition and core depletion, (2) studying operational and safety parameters evolution during \gls{MSBR} operation, (3) demonstrating that in a single-fluid two-region \gls{MSBR} conceptual design the undermoderated outer core zone II works as a virtual ``blanket", reduces neutron leakage and improves breeding ratio due to neutron energy spectral shift, and (4) determining the effect of fission product removal on the core neutronics.

The neutron population per cycle and the number of active/inactive cycles were chosen to obtain balance between reasonable uncertainty for a transport problem ($\leq$ 15 pcm for effective multiplication factor) and computational time. The \gls{MSBR} depletion and safety parameter computations were performed on 64 Blue Waters XK7 nodes (two AMD 6276 Interlagos CPU per node, 16 floating-point Bulldozer core units per node or 32 ``integer" cores per node, nominal clock speed is 2.45 GHz). The total computational time for achieving equilibrium composition was approximately 9,900 node-hours (18 core-years.)

\subsection{Effective multiplication factor}
Figure~\ref{fig:keff} demonstrates the effective multiplication factors obtained using SaltProc and SERPENT 2. The effective multiplication factors are calculated after removing fission products listed in Table~\ref{tab:reprocessing_list} and adding the fertile material at the end of ``cycle time"\footnote{The \gls{MSBR} program defined a ``cycle time" as the amount of time required to remove 100\% of a target nuclide from a fuel salt.} which was fixed at 3 days for this work. The effective multiplication factor fluctuates significantly as a result of the batch-wise nature of this online reprocessing strategy. 
\begin{figure}[hbp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{keff.png}
  \caption{Effective multiplication factor dynamics for full-core \gls{MSBR} model for a 60-year reactor operation. The confidence interval $\sigma=\pm15pcm$ is shaded.}
  \label{fig:keff}
\end{figure}
\FloatBarrier
First, SERPENT calculates the effective multiplication factor for the beginning of cycle time (fresh fuel composition for the first step). Next, it computes the new fuel salt composition for the end of a 3-day depletion step. The corresponding effective multiplication factor is much smaller than the previous one. Finally, SERPENT calculates $k_{eff}$ for the depleted composition after applying feeds and removals, and this increases accordingly since major reactor poisons (e.g. Xe, Kr) are removed, while fresh fissile material ($^{233}$U) from the protactinium decay tank is added.

Additionally, the presence of rubidium, strontium, cesium, and barium in the core are disadvantageous to reactor physics. In fact, removal of these elements every 3435 days causes the multiplication factor to jump by approximately 450 pcm, and limits using the batch approach for online reprocessing simulation. Overall, the effective multiplication factor gradually decreases from 1.075 to $k_{eff} \approx 1.02$ at equilibrium after approximately 6 years of irradiation.

The analysis of the fuel salt composition evolution provides more comprehensive information about the equilibrium state. Figure~\ref{fig:adens_eq} shows major nuclides which have a strong influence on the reactor core physics normalized separately for each isotope by average atomic density, at the beginning of each depletion time step. Concentration of $^{233}$U, $^{232}$Th, $^{233}$Pa, and $^{232}$Pa in fuel salt change insignificantly after approximately 2500 days of operation. Particularly, $^{233}$U number density fluctuates less than 0.8\% in the time interval from 16 to 20 years of operation, hence,a quasi-equlibrium state was achieved after 16 years of reactor operation.
\begin{figure}[htp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{major_isotopes_adens.png}
  \caption{Number density of major nuclides during 60 years of reactor operation.}
  \label{fig:adens_eq}
\end{figure}
In contrast, a wide variety of nuclides, including fissile isotopes (e.g. $^{235}$U) and non-fissile strong absorbers (e.g. $^{234}$U), keep accumulating in the core. Figure~\ref{fig:fissile_short} demonstrate production of fissile isotopes in the core. In the end of considered operational time the core contains significant $^{235}$U ($\approx10^{-5}$ atom/b-cm), $^{239}$Pu ($\approx5\times10^{-7}$ atom/b-cm), and $^{241}$Pu ($\approx 5\times10^{-7}$ atom/b-cm). Meanwhile, the equilibrium number density of the target fissile isotope $^{233}$U was approximately 7.97$\times10^{-5}$ atom/b-cm. Thus, production of new fissile materials in the core as well as $^{233}$U breeding make it possible to compensate for negative effects of strong absorber accumulation and keep the reactor critical.
\begin{figure}[htp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{fissile_short.png}
  \caption{Number density of fissile in epithermal spectrum nuclides accumulation during the reactor operation.}
  \label{fig:fissile_short}
\end{figure}
\subsection{Neutron spectrum}
Figure~\ref{fig:spectrum} shows the normalized neutron flux spectrum for the full-core \gls{MSBR} model in the energy range from $10^{-8}$ to $10$ MeV. The neutron energy spectrum at equilibrium is harder than at startup due to $^{238}$Pu, $^{239}$Pu, $^{240}$Pu, $^{241}$Pu, and $^{242}$Pu accumulation in the core during reactor operation. 
\begin{figure}[t!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{spectrum.png} 
  \caption{Neutron flux energy spectrum normalized by unit lethargy for initial and equilibrium fuel salt composition.}
  \label{fig:spectrum}
\end{figure}

Figure~\ref{fig:spectrum_zones} shows that zone I produced much more thermal neutrons than zone II, indicating that the majority of fissions occured in the central part of the core. In the undermoderated zone II, the neutron energy spectrum is harder which leads to more capture of neutrons by $^{232}$Th and helps a achieve relatively high breeding ratio. Moreover, the (n,$\gamma$) resonance energy range in $^{232}$Th is from 10$^{-4}$ to 10$^{-2}$ MeV. Therefore, the moderator-to-fuel ratio for zone II was chosen to shift the neutron energy spectrum in this range. Furthermore, in the central core region (zone I), the neutron energy spectrum shifts to a harder spectrum over 20 years of reactor operation. In contrast, in the outer core region (zone II) a similar spectral shift takes place at a reduced scale. This resuls is in a good agreement with original ORNL report \cite{robertson_conceptual_1971} and most recent whole-core steady-state study \cite{park_whole_2015}.

It is important to obtain the epithermal and thermal spectra to produce $^{233}$U from $^{232}$Th because the radiative capture cross section of thorium decreases monotonically from $10^{-10}$ MeV to $10^{-5}$ MeV. Hardening the spectrum tends to significantly increase resonance absorption in thorium and decrease the absorptions in fissile and construction materials. Thus, a signficant amount fissile material will be needed to make the reactor critical.

\begin{figure}[htp!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{spectrum_zones.png} 
  \caption{Neutron flux energy spectrum in different core regions normalized by unit lethargy for the initial and equilibrium fuel salt composition.}
  \label{fig:spectrum_zones}
\end{figure}

\subsection{Neutron flux}
Figure~\ref{fig:radial_flux} shows the radial distribution of fast and thermal neutron flux for both initial and equilibrium composition. The neutron flux has the same shape for both compositions but the equilibrium case has a harder spectrum. A significant spectral shift was observed for the central region of the core (zone I) when for the outer region (zone II) it is negligible for fast but notable for thermal neutrons. This neutron flux radial distribution is in a good agreement with original ORNL report \cite{robertson_conceptual_1971}. Overall, spectrum hardening during \gls{MSBR} operation should be carefully studied for designing the reactivity control system.

\begin{figure}[htp!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{radial_flux.png} 
  \caption{Radial neutron flux distribution for initial and equilibrium fuel salt composition.}
  \label{fig:radial_flux}
\end{figure}
\FloatBarrier

\subsection{Power and breeding distribution}
Table~\ref{tab:powgen_fraction} shows the power fraction in each zone for initial and equilibrium fuel composition. Figure~\ref{fig:pow_den} demonstrates the normalized power distribution of the \gls{MSBR} quater core at both states. For both the initial and equilibrium compositions, fission primarly occurs in the center of the core, namely zone I. The spectral shift during reactor operation results in different power fractions at startup and equilibrium, but most of the power is still generated in zone I. Figure~\ref{fig:breeding_den} shows the neutron capture reaction rate distribution for $^{232}$Th normalized by the total neutron flux for initial and equilibrium states. The distribution reflects the spatial distribution of $^{233}$Th production in the core. The thorium-232 then $\beta$-decays to $^{233}$Pa which is the precursor for $^{233}$U production. Accordingly, this characteristic represents the breeding distribution in the \gls{MSBR} core. Spectral shift does not cause significant changes in power nor in breeding distribution. Even after 20 years of operation, most of the power still is generated in zone I though the majority of $^{233}$Th is produced in zone II, which is in a good agreement with original ORNL report \cite{robertson_conceptual_1971}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
  \centering
  \caption{Power generation fraction in each zone for initial and equilibrium state.}
\begin{tabularx}{\textwidth}{ m | s | s } \hline
Core region      & Initial      & Equilibrium   \\   
Zone I           & 97.91\%      & 98.12\%   \\
Zone II          & 2.09\%       & 1.88\%   \\
\end{tabularx}
  \label{tab:powgen_fraction}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[htp!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{power_distribution.png} 
  \caption{Normalized power density for initial (top) and equilibrium (bottom) fuel salt composition.}
  \label{fig:pow_den}
\end{figure}
\FloatBarrier

\begin{figure}[htp!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{breeding_distribution.png} 
  \caption{$^{232}$Th neutron capture reaction rate normalized by total flux for initial (top) and equilibrium (bottom) fuel salt composition.}
  \label{fig:breeding_den}
\end{figure}
\FloatBarrier

\subsection{Temperature coefficient of reactivity}
Table~\ref{tab:tcoef} summarizes temperature effects on reactivity calculated in this work for both initial and equilibrium fuel composition, and compared with original \gls{ORNL} report data \cite{robertson_conceptual_1971}. Uncertainty for each temperature coefficient also appears in Table~\ref{tab:tcoef}. The main physical principle underlying the reactor temperature feedback is an expansion of matterial that is heated. When the fuel salt temperature increases, the density of the salt decreases, but at the same time, the total volume of fuel salt in the core remains constant because it is bounded by the graphite. When the graphite temperature increases, the density of graphite decreases creating additional space for fuel salt. To determine temperature coefficients, the cross section temperatures for fuel and moderator were changed from 900K to 1000K. Three different cases were considered:
\begin{enumerate}
  \item Temperature of fuel salt rising from 900K to 1000K.
  \item Temperature of graphite rising from 900K to 1000K.
  \item Whole reactor temperature rising from 900K to 1000K.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
  \centering
  \caption{Temperature coefficients of reactivity for initial and equilibrium state.}
\begin{tabularx}{\textwidth}{ m | s | s | s } \hline
   Reactivity coefficient [pcm/K]  & Initial      & Equilibrium  & Reference \cite{robertson_conceptual_1971} \\  
Fuel salt        & $-3.22\pm0.044$ & $-1.53\pm0.046$ & $-3.22$  \\
Moderator        & $+1.61\pm0.044$ & $+0.97\pm0.046$ & $+2.35$  \\
Total            & $-3.1\pm0.04$   & $-0.97\pm0.046$ & $-0.87$  \\
\end{tabularx}
  \label{tab:tcoef}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the first case, changes in the fuel temperature only impact fuel density. In this case, the geometry is unchanged because the fuel is a liquid. However, when the moderator heats up, both the density and the geometry change due to thermal expansion of the solid graphite blocks and reflector. Accordingly, the new graphite density was calculated using a linear temperature expansion coefficient of 1.3$\times10^{-6}$1/K \cite{robertson_conceptual_1971}. A new geometry input was created based on this information.

The fuel temperature coefficient (FTC) is negative for both initial and equilibrium fuel compositon due to thermal Doppler broadening of the resonance capture cross sections in the thorium and is in a good agreement with earlier research \cite{robertson_conceptual_1971,park_whole_2015}. The moderator temperature coefficient (MTC) is positive for startup composition and decreases during reactor operation because of spectrum hardening with fuel depletion. Finally, the total temperature coefficient of reactivity is negative for both cases, but decreases during reactor operation due to spectral shift. In summary, even after 20 years of operation the total temperature coefficient of reactivity is relatively large and negative during reactor operation, despite positive MTC, and affords excellent reactor stability and control.

\subsection{Reactivity control system rod worth}
Table~\ref{tab:rod_worth} summarizes the reactivity control system worth. During normal operation the control (graphite) rods are fully inserted, and the safety (B$_4$C) rods are fully withdrawn. To insert negative reactivity into the core, the graphite rods are gradually withdrawn from the core. In an accident, the safety rods would fall down into the core. The integral rod worths were calculated for various positions to separately estimate control (graphite) rod, safety (B$_4$C) rod, and the whole reactivity control system worth. Control rod integral worth is approximately 28 cents and stays almost constant during reactor operation. The safety rod integral worth decreases by  16.2\% during 20 years of operation because of neutron spectrum hardening and absorber accumulation in proximity to reactivity control system rods. This 16\% decline in control system worth should be taken into account in \gls{MSBR} accident analysis and safety justification.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[hb!]
  \centering
  \caption{Control system rod worth for initial and equilibrium fuel composition.}
\begin{tabularx}{\textwidth}{ m | s | s } \hline
\qquad\qquad Reactivity parameter  & \quad Initial      & \enspace Equilibrium      \\
Control (graphite) rod integral worth (cents)               & $\ 28.2\pm0.8$     & $\ 29.0\pm0.8$ \\ 
Safety (B$_4$C) rod integral worth (cents)                  & $251.8\pm0.8$    & $211.0\pm0.8$  \\
Total reactivity control system worth (cents)               & $505.8\pm0.7$    & $424.9\pm0.8$ \\
\end{tabularx}
  \label{tab:rod_worth}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Six Factor Analysis}
The effective multiplication factor could be expressed using formula:
\begin{align*}
k_{eff} = k_{inf} P_f  P_t \\
 = \eta \epsilon p f P_f P_t
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
  \centering
  \caption{Six factors for the full-core \gls{MSBR} model for initial and equilibrium fuel composition.}
\begin{tabularx}{\textwidth}{ m | s | s } \hline
	   \qquad\qquad\qquad Factors  & \qquad Initial      & \qquad Equilibrium   \\ 
Neutron reproduction factor ($\eta$)     & $1.3960\pm.000052$     & $1.3778\pm.00005$ \\ 
Thermal utilization factor (f)           & $0.9670\pm.000011$     & $0.9706\pm.00001$ \\
Resonance escape probability (p)         & $0.6044\pm.000039$     & $0.5761\pm.00004$ \\
Fast fission factor ($\epsilon$)         & $1.3421\pm.000040$     & $1.3609\pm.00004$ \\
Fast non-leakage probability (P$_f$)     & $0.9999\pm.000004$     & $0.9999\pm.000004$ \\
Thermal non-leakage probability (P$_t$)  & $0.9894\pm.000005$     & $0.9912\pm.00005$ \\ 
\end{tabularx}
  \label{tab:six_factor}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Table~\ref{tab:six_factor} summarizes the six factors for both initial and equilibrium fuel salt composition. The non-leakage probability for both fast and thermal neutrons does not change during reactor operation because these values are not largely affected by the neutron spectrum shift. In contrast, neutron reproduction factor ($\eta$), resonance escape probability (p), and fast fission factor ($\epsilon$) are considerably different between startup and equilibrium. As indicated in Figure~\ref{fig:spectrum} the neutron spectrum is softer for the initial state. Neutron spectrum hardening causes the fast fission to increase through the core lifetime. The opposite is true for the resonance escape probability. Finally, the neutron reproduction factor decreases during reactor operation due to accumulation of fissile plutonium isotopes.

\subsection{Thorium refill rate}
As was mentioned in Methods the only external feed material flow for this \gls{MSBR} reprocessing scheme is $^{232}$Th. Figure~\ref{fig:th_refill} shows the $^{232}$Th feed rate calculated for 20 years of reactor operation. Figure~\ref{fig:th_refill_spike} shows the large spikes up to 36 kg/day in a thorium consumption every 3435 days. This is required due to batch-wise removal of strong absorbers (Rb, Sr, Cs, Ba). The corresponding effective multiplication factor increase (Figure~\ref{fig:keff}) and breeding intensification leads to additional $^{232}$Th consumption. As indicated in Figure~\ref{fig:th_refill}, the average thorium feed rate increases during the first 500 days of operation and than steadily decreases due to spectrum hardening and accumulation of absorbers in the core. As a result, the average $^{232}$Th feed rate over 20 years of operation is about 2.39 kg/day. This results is in a good agreement with a recent online reprocessing study by \gls{ORNL} \cite{betzler_molten_2017} which reported thorium-232 refill rate for single-cell online reprocessing of 2.45 kg/day.

\begin{figure}[htp!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{Th_refill_rate.png} 
  \caption{$^{232}$Th feed rate over 20 years of \gls{MSBR} operation.}
  \label{fig:th_refill}
\end{figure}
\begin{figure}[htp!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{Th_refill_rate_spike.png} 
  \caption{Typical $^{232}$Th feed rate spike caused by strong absorbers (Rb, Sr, Cs, Ba) removal.}
  \label{fig:th_refill_spike}
\end{figure}
\FloatBarrier
